\bf Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2024-07-18 paper.qmd
Update texts of MNIST data set

The MNIST dataset consists of grayscale images of handwritten digits [@lecun2010]. @yingfan2021 used this dataset to demonstrate how PaCMAP preserves local structure. We selected the \gD{} embedding of PaCMAP for the handwritten digit 1 to assess whether this is a reasonable representation using our method. As shown in @fig-pacmap-author-img, the angle of the digit 1 images varies along the \gD{} structure.

@@

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->
```{r}
#| fig-cap: "$2\\text{-}D$ layout from PaCMAP applied for the digit 1 of the MNIST dataset with a specific parameter settings. The angle of the digit 1 varies along this structure. Images at the top-left of the $2\\text{-}D$ layout show the digit 1 angled more to the left, while images at the bottom-right show the digit 1 angled more to the right. Is this the best representation of the digit 1?"
#| label: fig-pacmap-author
#| fig-pos: H
#| out-height: 22%

@@

<!--Render the images in different selected location in 2D layout-->
```{r}
## Data with pixel values
mnist_data <- read_rds("data/mnist/mnist_digit_1.rds")

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c(3710, 2391, 2385, 5030, 6475,
                         7679, 3568, 1312, 3311, 1097,
                         3552, 7853, 6489, 7689, 6690,
                        1380, 6057, 2347, 5946, 3355,
                        4175, 3997, 5378, 387, 1854,
                        614, 3079, 1762, 5239, 3723,
                        5748, 728, 7419, 7794, 6233,
                        33, 2485, 5998, 318, 1761,
                        5690, 165, 517, 6935, 2682,
                        4962, 2264, 5563, 6369, 559,
                        5188, 4849, 2666, 3448, 3055,
                        3120, 6869, 6345, 4470, 7147)) |>
  mutate(instance.labs =  case_when(
    instance %in% c(3710, 2391, 2385, 5030, 6475) ~ "1",
    instance %in% c(7679, 3568, 1312, 3311, 1097) ~ "2",
    instance %in% c(3552, 7853, 6489, 7689, 6690) ~ "3",
    instance %in% c(1380, 6057, 2347, 5946, 3355) ~ "4",
    instance %in% c(4175, 3997, 5378, 387, 1854) ~ "5",
    instance %in% c(614, 3079, 1762, 5239, 372) ~ "6",
    instance %in% c(5748, 728, 7419, 7794, 6233) ~ "7",
    instance %in% c(5690, 165, 517, 6935, 2682) ~ "8",
    instance %in% c(4962, 2264, 5563, 6369, 559) ~ "9",
    instance %in% c(5188, 4849, 2666, 3448, 3055) ~ "10",
    instance %in% c(3120, 6869, 6345, 4470, 7147) ~ "11",
    .default = "12"
    ))

## To generate labels
instance.labs <- as.character(rep(1:12, each = 5))

img_sample <- pixels_gathered |> 
  ggplot(aes(x, y, fill = value)) +  
  geom_tile() +
  facet_wrap(~ factor(instance, levels = c(3710, 2391, 2385, 5030, 6475,
                         7679, 3568, 1312, 3311, 1097,
                         3552, 7853, 6489, 7689, 6690,
                        1380, 6057, 2347, 5946, 3355,
                        4175, 3997, 5378, 387, 1854,
                        614, 3079, 1762, 5239, 3723,
                        5748, 728, 7419, 7794, 6233,
                        33, 2485, 5998, 318, 1761,
                        5690, 165, 517, 6935, 2682,
                        4962, 2264, 5563, 6369, 559,
                        5188, 4849, 2666, 3448, 3055,
                        3120, 6869, 6345, 4470, 7147)), nrow = 5) +
  coord_fixed() +
  scale_fill_continuous_sequential(palette = "Grays") +
  theme(strip.background = element_blank(),
        strip.text.x = element_blank(), 
        legend.position = "none")
```

```{r}
#| fig-cap: "Images of the handwritten digit 1 are ordered from the bottom-right to the top-left of the $2\\text{-}D$ structure. The angle of the digit varies along this structure. Images at the bottom-right of the $2\\text{-}D$ layout show the digit 1 angled more to the right, while images at the top-left show the digit 1 angled more to the left. This demonstrates how the angle changes from right to left along the $2\\text{-}D$ structure."
#| label: fig-pacmap-author-img
#| fig-pos: H
#| out-height: 50%

img_sample
```

@@

According to @fig-mnist1-sc1, the non-linear continuous structure observed in the \gD{} representation of PaCMAP (see @fig-pacmap-author) is also visible when visualizing the model overlaid on the data space. This indicates that PaCMAP accurately captures the structure of the \pD{} data. Additionally, the model shows a twisted pattern within the non-linear structure in \pD{} space (see @fig-mnist1-sc2), which is an additional pattern not visible in the \gD{} representation (see @fig-pacmap-author). Furthermore, as shown in @fig-mnist1-sc3, some long edges exist in the \pD{} space that are not recognized as long edges in the \gD{} representation. However, PaCMAP is a reasonable \gD{} representation of MNIST digit 1 data. Because PaCMAP preserves the local structure.

@@

There are certain data points that exhibit high error rates due to their deviation from the usual \pD{} data structure, which makes them anomalies (see @fig-model-mnist (b)). These anomalies can be classified into two types: those that are anomalies within the non-linear structure and those that lie outside of it. The images associated with high model error points within the non-linear structure display different patterns of the digit 1, as shown in @fig-mnist-anomalies (a). However, when comparing these images to the ones found outside of the non-linear structure, it becomes evident that the latter display different patterns of the digit 1 (see @fig-mnist-anomalies (b)).

<!--images that occur large error-->
<!-- within the nonlinear structure-->
```{r}
pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c( 4699, 4785, 5173, 6671, 2186, 4636, 4882, 
                          773, 2825, 6169, 6948, 5987, 1282, 7715, 
                          622, 2989, 6912, 4370, 1227, 4042, 2444, 
                          836, 1654))

>>>>

The MNIST dataset consists of grayscale images of handwritten digits [@lecun2010]. @yingfan2021 used this dataset to demonstrate how PaCMAP preserves non-linear structure in \pD{}. To evaluate whether PaCMAP provides a reasonable representation of the data, the \gD{} embedding of the handwritten digit 1 was selected. As shown in @fig-pacmap-author, the angle of the digit 1 images varies along the \gD{} structure.

@@

<!--PaCMAP param: n_components=2, n_neighbors=10, init=random, MN_ratio=0.9, FP_ratio=2.0-->
```{r}
#| fig-cap: "$2\\text{-}D$ layout from PaCMAP applied for the digit 1 of the MNIST dataset with a specific parameter settings. The angle of the digit 1 varies along this structure. Images at the top-left of the $2\\text{-}D$ layout show the digit 1 angled more to the left, while images at the bottom-right show the digit 1 angled more to the right. Is this a reasonable representation of the digit 1?"
#| label: fig-pacmap-author
#| fig-pos: H
#| out-height: 22%

@@

@@

According to @fig-mnist1-sc1, the non-linear continuous structure observed in the \gD{} representation of PaCMAP (@fig-pacmap-author) is also visible when visualizing the model overlaid on the data space. This indicates that PaCMAP accurately captures the structure of the \pD{} data. Additionally, the model shows a twisted pattern within the non-linear structure in \pD{} space (@fig-mnist1-sc2), which is an additional pattern not visible in the \gD{} representation (@fig-pacmap-author). Furthermore, as shown in @fig-mnist1-sc3, some long edges exist in the \pD{} space that are not recognized as long edges in the \gD{} representation. However, PaCMAP is a reasonable \gD{} representation of MNIST digit 1 data, because PaCMAP preserves the non-linear structure present in the \pD{} data.

@@

There are certain data points that exhibit high error rates due to their deviation from the usual \pD{} data structure, which makes them anomalies (@fig-model-mnist (b)). These anomalies can be classified into two types: those that are anomalies within the non-linear structure and those that lie outside of it. The images associated with high model error points within the non-linear structure display different patterns of the digit 1, as shown in @fig-mnist-anomalies (a). However, when comparing these images to the ones found outside of the non-linear structure, it becomes evident that the latter display different patterns of the digit 1 (@fig-mnist-anomalies (b)).

<!--images that occur large error-->
<!-- within the nonlinear structure-->
```{r}
## Data with pixel values
mnist_data <- read_rds("data/mnist/mnist_digit_1.rds")

pixels_gathered <-  mnist_data |>
  mutate(instance = row_number()) |>
  gather(pixel, value, -Label, -instance) |>
  tidyr::extract(pixel, "pixel", "(\\d+)", convert = TRUE) |>
  mutate(pixel = pixel - 2, x = pixel %% 28, y = 28 - pixel %/% 28) |>
  filter(instance %in% c( 4699, 4785, 5173, 6671, 2186, 4636, 4882, 
                          773, 2825, 6169, 6948, 5987, 1282, 7715, 
                          622, 2989, 6912, 4370, 1227, 4042, 2444, 
                          836, 1654))

### 2024-07-18 paper.tex
Update texts of MNIST data set

The MNIST dataset consists of grayscale images of handwritten digits
\citep{lecun2010}. \citet{yingfan2021} used this dataset to demonstrate
how PaCMAP preserves local structure. We selected the \gD{} embedding of
PaCMAP for the handwritten digit 1 to assess whether this is a
reasonable representation using our method. As shown in
Figure~\ref{fig-pacmap-author-img}, the angle of the digit 1 images
varies along the \gD{} structure.

@@

\caption{\label{fig-pacmap-author}\(2\text{-}D\) layout from PaCMAP
applied for the digit 1 of the MNIST dataset with a specific parameter
settings. The angle of the digit 1 varies along this structure. Images
at the top-left of the \(2\text{-}D\) layout show the digit 1 angled
more to the left, while images at the bottom-right show the digit 1
angled more to the right. Is this the best representation of the digit
1?}

\includegraphics[width=1\textwidth,height=0.5\textheight]{paper_files/figure-pdf/fig-pacmap-author-img-1.pdf}

\caption{\label{fig-pacmap-author-img}Images of the handwritten digit 1
are ordered from the bottom-right to the top-left of the \(2\text{-}D\)
structure. The angle of the digit varies along this structure. Images at
the bottom-right of the \(2\text{-}D\) layout show the digit 1 angled
more to the right, while images at the top-left show the digit 1 angled
more to the left. This demonstrates how the angle changes from right to
left along the \(2\text{-}D\) structure.}

@@

According to Figure~\ref{fig-mnist1-sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP (see
Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \pD{} data. Additionally, the model shows
a twisted pattern within the non-linear structure in \pD{} space (see
Figure~\ref{fig-mnist1-sc2}), which is an additional pattern not visible
in the \gD{} representation (see Figure~\ref{fig-pacmap-author}).
Furthermore, as shown in Figure~\ref{fig-mnist1-sc3}, some long edges
exist in the \pD{} space that are not recognized as long edges in the
\gD{} representation. However, PaCMAP is a reasonable \gD{}
representation of MNIST digit 1 data. Because PaCMAP preserves the local
structure.

@@

There are certain data points that exhibit high error rates due to their
deviation from the usual \pD{} data structure, which makes them
anomalies (see Figure~\ref{fig-model-mnist} (b)). These anomalies can be
classified into two types: those that are anomalies within the
non-linear structure and those that lie outside of it. The images
associated with high model error points within the non-linear structure
display different patterns of the digit 1, as shown in
Figure~\ref{fig-mnist-anomalies} (a). However, when comparing these
images to the ones found outside of the non-linear structure, it becomes
evident that the latter display different patterns of the digit 1 (see
Figure~\ref{fig-mnist-anomalies} (b)).

>>>>

The MNIST dataset consists of grayscale images of handwritten digits
\citep{lecun2010}. \citet{yingfan2021} used this dataset to demonstrate
how PaCMAP preserves non-linear structure in \pD{}. To evaluate whether
PaCMAP provides a reasonable representation of the data, the \gD{}
embedding of the handwritten digit 1 was selected. As shown in
Figure~\ref{fig-pacmap-author}, the angle of the digit 1 images varies
along the \gD{} structure.

@@

\caption{\label{fig-pacmap-author}\(2\text{-}D\) layout from PaCMAP
applied for the digit 1 of the MNIST dataset with a specific parameter
settings. The angle of the digit 1 varies along this structure. Images
at the top-left of the \(2\text{-}D\) layout show the digit 1 angled
more to the left, while images at the bottom-right show the digit 1
angled more to the right. Is this a reasonable representation of the
digit 1?}

@@

According to Figure~\ref{fig-mnist1-sc1}, the non-linear continuous
structure observed in the \gD{} representation of PaCMAP
(Figure~\ref{fig-pacmap-author}) is also visible when visualizing the
model overlaid on the data space. This indicates that PaCMAP accurately
captures the structure of the \pD{} data. Additionally, the model shows
a twisted pattern within the non-linear structure in \pD{} space
(Figure~\ref{fig-mnist1-sc2}), which is an additional pattern not
visible in the \gD{} representation (Figure~\ref{fig-pacmap-author}).
Furthermore, as shown in Figure~\ref{fig-mnist1-sc3}, some long edges
exist in the \pD{} space that are not recognized as long edges in the
\gD{} representation. However, PaCMAP is a reasonable \gD{}
representation of MNIST digit 1 data, because PaCMAP preserves the
non-linear structure present in the \pD{} data.

@@

There are certain data points that exhibit high error rates due to their
deviation from the usual \pD{} data structure, which makes them
anomalies (Figure~\ref{fig-model-mnist} (b)). These anomalies can be
classified into two types: those that are anomalies within the
non-linear structure and those that lie outside of it. The images
associated with high model error points within the non-linear structure
display different patterns of the digit 1, as shown in
Figure~\ref{fig-mnist-anomalies} (a). However, when comparing these
images to the ones found outside of the non-linear structure, it becomes
evident that the latter display different patterns of the digit 1
(Figure~\ref{fig-mnist-anomalies} (b)).
